---
title: "Class 12 RNA SeqGalaxy A16246401"
author: "Scott MacLeod"
format: pdf
---

## Class 12 RNA SeqGalaxy, Proportion of G/G in a Population

Downloaded CSV file from Ensemble

We are going to read the CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```


Now we are going to look at a GBR population with 91 individuals.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```
```{r}
table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100
```

This variant that is associated with childhood asthma is more frequent in the GBR population that the MXL population.

Now let's dig into this further. 



# Section 4 (Homework)

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
```
>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

The boxplot shows the median expression for each of the genotypes. For the allele A/A the median expression is around 31, while A/G is 25, and G/G is 20. 

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

There is a statistically significant difference in expression between the two homozygous genotypes of A/A and G/G. This means that the SNP does effet the expression of ORMDL3.

